Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-17 paper.tex
Update Jackson et al breakpoints for sc2ts

\begin{table} \centering
\begin{tabular}{lllll}
\toprule
Group        & Sequences & Method & Interval & Parents \\
\midrule
A (XA)       & 4   & Jackson        &  21,256--21,613 & B.1.177/B.1.1.7 \\
             &     &\texttt{sc2ts} &  21,256--22,227 & B.1.177.18/B.1.1.7 \\
\cmidrule{3-5}
B            & 2   & Jackson        &  6,529--6,953 & B.1.36/B.1.1.7  \\
             &     &\texttt{sc2ts} &  6,529--6,954 & B.1.36/B.1.1.7  \\
\cmidrule{3-5}
C            & 3   &Jackson        &  25,997--27,441 &  B.1.1.7/B.1.221 \\
             &     & \texttt{sc2ts} &  25,997--27,972 &  B.1.1.7/B.1.221 \\
\cmidrule{3-5}
D            & 3   & Jackson        &  21,576--23,063 &  B.1.36.17/B.1.1.7 \\
             &     & \texttt{sc2ts} &  22,445--23,063 &  B.1.36.39/B.1.1.7 \\
\bottomrule
\end{tabular}
\caption{\label{tab:jackson}Comparison of recombination breakpoint
intervals and parent lineages for Groups A-D
reported by \cite{Jackson2021-ik} with the corresponding
recombination events in the Wide ARG.
The second column gives the number of sequences in the group.
Intervals are given in left-inclusive, right-exclusive form.
Details for all 16 sequences are
given in Table~\ref{tab:jackson_supplement}.}
\end{table}

@@

In this section we focus on the detailed genealogy of samples that have been
previously identified as recombinants, i.e. given a Pango designation that
starts with an ``X''. Since the wide ARG is restricted to data collected prior
to mid-2021, it contains samples from only three Pango-designated recombinant
lineages: XA, XB, and XC. To investigate the diversity of genealogies
associated with recombinants, we therefore devote more analysis to patterns in
the long ARG. As in previous sections, we exclude singleton recombinants where
possible.

\subsubsection{Wide ARG}
%% Quick overview of the wide ARG.
In the wide ARG, 44
samples are designated as XA by both Nextclade and GISAID, 237 are designated
as XB by Nextclade (231 by GISAID), and 6 as XC by Nextclade (none by GISAID).
After removing singleton recombinants, XA numbers remain unchanged, but XB is
reduced to 235 (229) and XC to 4 (0). We correctly identify that all samples
designated as XA, XB, or XC by any method have one or more recombination nodes
in their ancestry.

@@

For XC, the wide ARG identifies more than one originating recombination node
(see "multiple origins" below). However, as none of the samples designated as
XC by Nextclade are designated as XC by GISAID, these patterns could be due to
mistaken labelling, hence we do not examine XC further.

\subsubsection{Long ARG}
%% high level summary of the pango recombinant lineages and samples in the Long ARG
The long ARG (subsampled to mid-2022) contains 33
Nextclade-designated Pango recombinant lineages (XA to XAK; Supplementary Table
XXX) comprising 749 samples (711 when singleton recombinants are removed). In
contrast, GISAID designates 38 Pango recombinant lineages in the long ARG,
comprising only 515 samples. In the analyses below, we mainly focus on the
larger number of Nextclade designated samples.

@@

In total out of the 31 Pango recombinant lineages in the long ARG we identify 16 as
having a single origin. Supplementary Table~\ref{tab:pango-single-origin} summarises these
together with their inferred parents, showing good agreement between \texttt{sc2ts} and
the official Pango designated parents, with \texttt{sc2ts} providing more a precise
parent designation. Six of the 16 lineages (XA, XAC, XAE, XF, XK, and XS) are of the
simple-origin (XA) type, whereas the rest are of the XAG type, and include multiple Pango
designated lineages as descendants of the originating recombination event. In some cases,
these may be, however, be a result of erroneous Pango designations.

@@

% FIXME there's too much vspace between the rows here. Probably best to redo
% the formatting.
\begin{table} \centering
\begin{tabular}{l|c|c|c|c|c} \hline
\multicolumn{1}{c}{} & \multicolumn{3}{c}{\textbf{Jackson et al. (2021)}} &
\multicolumn{2}{c}{\textbf{sc2ts (wide ARG)}} \\ \hline
\textbf{Sample} &
\textbf{Group} & \textbf{Parents} & \thead{Breakpoint \\
interval(s)} &
\textbf{Parents} & \thead{Breakpoint \\ interval(s)} \\
\hline ALDP-11CF93B & A
    & \thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} &
21,256–22,227 \\
    ALDP-125C4D7 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 &
    \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
ALDP-130BB95 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 &
    \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
LIVE-DFCFFE & A & \thead{B.1.177 \\ Alpha} & 18,998–20,294 &
    \thead{B.1.177.18 \\ Alpha} & 21,256–22,227 \\
QEUH-CCCB30 & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 &
    \thead{B.1.36 \\ Alpha} & 6,529–6,954 \\
QEUH-CD0F1F & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 &
    \thead{B.1.36 \\ Alpha} & 6,529–6,954 \\
MILK-1166F52 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 &
    \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
MILK-11C95A6 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 &
    \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
QEUH-109B25C & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 &
    \thead{Alpha \\ B.1.221} & 25,997–27,972 \\
MILK-126FE1F & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 &
    \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\
RAND-12671E1 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 &
    \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\
RAND-128FA33 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 &
    \thead{B.1.36.39 \\ Alpha} & 22,445–23,063 \\
CAMC-CBA018 & n/a & \thead{B.1.177 \\ Alpha} & 20,389–21,254 &
    \thead{B.1.177 \\ Alpha} & 16,177–21,255 \\
CAMC-CB7AB3 & n/a & \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3,267–4,474 \\ 20,389–21,254} &
     \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3,268–5,388 \\ 16,177–21,255} \\
MILK-103C712 & n/a & \thead{B.1.177.17 \\ Alpha} & \thead{408–444 \\ 26,801–27,876} &
    n/a & n/a \\
QEUH-1067DEF & n/a & \thead{Alpha \\ B.1.177.9} & 10,523–10,869 &
    \thead{Alpha \\ B.1.177} & 7,729–10,870 \\ \hline
\end{tabular}
\caption{\label{tab:jackson_supplement}Recombinant sequences involving the Alpha (B.1.1.7) variant reported by \cite{Jackson2021-ik} have recombinant ancestry in the wide ARG. The breakpoint intervals and Pango lineage assignments of the parents were taken from Table 2 (3SEQ results) of Jackson et al., except the Pango lineage assignment of the parents of group B recombinants, which were taken from Table 1 (motif-based results).}
\end{table}

@@

\begin{sidewaystable}
% \begin{footnotesize}
\centering
\begin{tabular}{p{1cm}p{1.2cm}p{4.2cm}lll}
\toprule
Focal Pango  & Num origins & Num focal samples (further split by origin, *=nested) & Official Pango parents & \textbf{Main clade}: sc2ts parents & \textbf{Main clade}: other descendants \\
\midrule
\bfseries XA & 1 & \textbf{5} & B.1.1.7 + B.1.177 & B.1.1.7 + B.1.177.18 &  \\
\bfseries XF & 1 & \textbf{2} & B.1.617.2* + BA.1* & AY.4 + BA.1 &  \\
\bfseries XG & 1 & \textbf{32} & BA.1* + BA.2* & BA.1.17 + BA.2 & XAB: 1/48 \\
\bfseries XH & 1 & \textbf{11} & BA.1* + BA.2* & BA.1.20 + BA.2.9 & XAF: 34/35, B.1.1.529: 2, XE: 3/163 \\
\bfseries XK & 1 & \textbf{3} & BA.1* + BA.2* & BA.1.1.1 + BA.2 &  \\
\bfseries XL & 1 & \textbf{10} & BA.1* + BA.2* & BA.1.17.2 + BA.2 & XAB: 1/48, XU: 1/3 \\
\bfseries XR & 1 & \textbf{8} & BA.1.1* + BA.2* & BA.1.1 + BA.2 & XQ: 9/12, XAB: 2/48 \\
\bfseries XS & 1 & \textbf{4} & B.1.617.2* + BA.1.1* & AY.36 + BA.1.1 &  \\
\bfseries XT & 1 & \textbf{1} & BA.1* + BA.2* & BA.2.23 + Unknown & BA.2.23: 1 \\
\bfseries XV & 1 & \textbf{1} & BA.1* + BA.2* & BA.1.1 + BA.2 + Unknown & BA.2: 8 \\
\bfseries XW & 1 & \textbf{11} & BA.1* + BA.2* & BA.1.1.15 + BA.2 & XN: 2/13 \\
\bfseries XY & 1 & \textbf{6} & BA.1* + BA.2* & BA.1.1 + BA.2 & XAF: 1/35 \\
\bfseries XAA & 1 & \textbf{8} & BA.1* + BA.2* & BA.1 + BA.2.9 & XAB: 38/48, XAG: 17/17, XU: 1/3, XQ: 2/12 \\
\bfseries XAC & 1 & \textbf{27} & BA.1* + BA.2* & BA.1.17.2 + BA.2.3 &  \\
\bfseries XAE & 1 & \textbf{9} & BA.1* + BA.2* & BA.1 + BA.2 &  \\
\bfseries XAG & 1 & \textbf{17} & BA.1* + BA.2* & BA.1 + BA.2.9 & XAB: 38/48, XAA: 8/8, XU: 1/3, XQ: 2/12 \\
\bfseries XB & 2 & 57 (\textbf{57}, 1)* & B.1.631 + B.1.634 & B.1 + B.1.627 & B.1.634: 3, B.1.631: 7 \\
\bfseries XC & 2 & 3 (1, \textbf{2}) & AY.29 + B.1.1.7 & AY.103 + B.1.1.7 &  \\
\bfseries XD & 2 & 4 (1, \textbf{3}) & B.1.617.2* + BA.1* & AY.4 + BA.1.15 &  \\
\bfseries XJ & 2 & 2 (1, 1) & BA.1* + BA.2* & No main clade & No main clade \\
\bfseries XAD & 2 & 5 (1, \textbf{4}) & BA.1* + BA.2* & BA.1.1 + BA.2 &  \\
\bfseries XAF & 2 & 35 (\textbf{34}, 1) & BA.1* + BA.2* & BA.1.20 + BA.2.9 & XH: 11/11, B.1.1.529: 2, XE: 3/163 \\
\bfseries XAH & 2 & 12 (\textbf{10}, 2) & BA.1* + BA.2* & BA.1 + BA.2.10 & XZ: 61/66, XAD: 1/5 \\
\bfseries XQ & 3 & 12 (2, \textbf{9}, 1) & BA.1.1* + BA.2* & BA.1.1 + BA.2 & XR: 8/8, XAB: 2/48 \\
\bfseries XU & 3 & 3 (1, 1, 1) & BA.1* + BA.2* & No main clade & No main clade \\
\bfseries XM & 4 & 47 (2, 1, \textbf{40}, 4) & BA.1.1* + BA.2* & BA.1.1 + BA.2 &  \\
\bfseries XN & 4 & 13 (2, 2, \textbf{7}, 2) & BA.1* + BA.2* & BA.1 + BA.2 &  \\
\bfseries XAJ & 4 & 5 (1, 1, \textbf{3}, 1)* & BA.2.12.1* + BA.4* & Unknown & BA.5: 1 \\
\bfseries XE & 5 & 163 (3, \textbf{155}, 1, 3, 2)* & BA.1* + BA.2* & BA.1.17.2 + BA.2 &  \\
\bfseries XZ & 5 & 66 (1, 1, \textbf{61}, 2, 1) & BA.1* + BA.2* & BA.1 + BA.2.10 & XAH: 10/12, XAD: 1/5 \\
\bfseries XAB & 8 & 48 (1, \textbf{38}, 1, 2, 2, 2, 18, 2)* & BA.1* + BA.2* & BA.1 + BA.2.9 & XAA: 8/8, XAG: 17/17, XU: 1/3, XQ: 2/12 \\
\bottomrule
\end{tabular}
\caption{\label{tab:pango-single-origin}
Summary of all Pango X-lineages in the long ARG. In cases of multiple origins, most X-lineages have a single ``main clade'' (in bold). 
Sc2ts inferred parents for the main clade are based on imputed NextClade Pango lineages. Official parents
are taken from Pango designation alias key (TODO: explain the asterisks). Other inferred samples within the main clade are listed, grouped by NextClade Pango designation and clade count / total ARG count (the latter being omitted for non-recombinant designations).
}
\end{sidewaystable}

\begin{sidewaystable}
\centering
\begin{tabular}{llp{1.2cm}llrl}
\toprule
Causal date & Sc2ts inferred parents & Causal lineage & Causal strain & Breakpoint & $n_{XE}$ & Other descendants \\
\midrule
2022-01-06 & BA.1.20 + BA.2.9 & XH & Denmark/DCGC-323778/2022 & (10448,11538) & 3 & XAF: 34/35, XH: 11/11, B.1.1.529: 2 \\
2022-02-11 & BA.1.17.2 + BA.2 & XE & England/MILK-3674F1A/2022 & (10448,11538) & 155 &  \\
2022-02-17 & Unknown + BA.2 & XE & Turkey/HSGM-G2220/2022 & (17411,19956) & 1 & B.1.1.529: 1 \\
2022-04-09 & BA.1 + BA.2 & XE & Belgium/ULG-25782/2022 & (9867,10199) & 3 &  \\
2022-04-20 & BA.1.1.15 + Unknown & XE & Canada/BC-BCCDC-424071/2022 & (10448,11538) & 2 &  \\
\bottomrule
\end{tabular}\caption{\label{tab:multiple_origins_table}
Summary of inferred multiple ``origins'' for Nextclade-designated XE samples
(Official: BA.1* \& BA.2, 10448–11287, defined from UK and Irish strains, earliest
on 2022/1/16;  see \protect\url{https://github.com/cov-lineages/pango-designation/issues/454}).
XE samples are grouped by their most recent ancestral recombination node;
Details are also provided for the ``causal strain'' that led to the recombination
being proposed by \texttt{sc2ts}.
}
\end{sidewaystable}

>>>>

\begin{table} \centering
\begin{tabular}{lllll}
\toprule
Group        & Sequences & Method & Interval & Parents \\
\midrule
A (XA)       & 4   & Jackson        &  21,256--21,613 & B.1.177/B.1.1.7 \\
             &     &\texttt{sc2ts} &  21,256--22,228 & B.1.177.18/B.1.1.7 \\
\cmidrule{3-5}
B            & 2   & Jackson        &  6,529--6,954 & B.1.36/B.1.1.7  \\
             &     &\texttt{sc2ts} &  6,529--6,955 & B.1.36/B.1.1.7  \\
\cmidrule{3-5}
C            & 3   &Jackson        &  25,997--27,441 &  B.1.1.7/B.1.221 \\
             &     & \texttt{sc2ts} &  25,997--27,973 &  B.1.1.7/B.1.221 \\
\cmidrule{3-5}
D            & 3   & Jackson        &  21,576--23,063 &  B.1.36.17/B.1.1.7 \\
             &     & \texttt{sc2ts} &  22,445--23,064 &  B.1.36.39/B.1.1.7 \\
\bottomrule
\end{tabular}
\caption{\label{tab:jackson}Comparison of recombination breakpoint
intervals and parent lineages for Groups A-D
reported by \cite{Jackson2021-ik} with the corresponding
recombination events in the Wide ARG.
The second column gives the number of sequences in the group.
Intervals are half-closed, so that for an interval $\ell$--$r$
the breakpoint $x$ can be $\ell \leq x < r$.
Details for all 16 sequences are
given in Table~\ref{tab:jackson_supplement}.}
\end{table}

@@

In this section we focus on the detailed genealogy of samples that have been
previously identified as recombinants, i.e. given a Pango designation that
starts with an ``X''. Since the Wide ARG is restricted to data collected prior
to mid-2021, it contains samples from only three Pango-designated recombinant
lineages: XA, XB, and XC. To investigate the diversity of genealogies
associated with recombinants, we therefore devote more analysis to patterns in
the Long ARG. As in previous sections, we exclude singleton recombinants where
possible.

\subsubsection{Wide ARG}
%% Quick overview of the wide ARG.
In the Wide ARG, 44
samples are designated as XA by both Nextclade and GISAID, 237 are designated
as XB by Nextclade (231 by GISAID), and 6 as XC by Nextclade (none by GISAID).
After removing singleton recombinants, XA numbers remain unchanged, but XB is
reduced to 235 (229) and XC to 4 (0). We correctly identify that all samples
designated as XA, XB, or XC by any method have one or more recombination nodes
in their ancestry.

@@

For XC, the Wide ARG identifies more than one originating recombination node
(see "multiple origins" below). However, as none of the samples designated as
XC by Nextclade are designated as XC by GISAID, these patterns could be due to
mistaken labelling, hence we do not examine XC further.

\subsubsection{Long ARG}
%% high level summary of the pango recombinant lineages and samples in the Long ARG
The Long ARG (subsampled to mid-2022) contains 33
Nextclade-designated Pango recombinant lineages comprising 749 samples (711 when singleton recombinants are removed). In
contrast, GISAID designates 38 Pango recombinant lineages in the long ARG,
comprising only 515 samples. In the analyses below, we mainly focus on the
larger number of Nextclade designated samples.

@@

In total out of the 31 Pango recombinant lineages in the Long ARG we identify 16 as
having a single origin. Supplementary Table~\ref{tab:pango-single-origin} summarises these
together with their inferred parents, showing good agreement between \texttt{sc2ts} and
the official Pango designated parents, with \texttt{sc2ts} providing more a precise
parent designation. Six of the 16 lineages (XA, XAC, XAE, XF, XK, and XS) are of the
simple-origin (XA) type, whereas the rest are of the XAG type, and include multiple Pango
designated lineages as descendants of the originating recombination event. In some cases,
these may be, however, be a result of erroneous Pango designations.

@@

% FIXME there's too much vspace between the rows here. Probably best to redo
% the formatting.
\begin{table} \centering
\begin{tabular}{l|c|c|c|c|c} \hline
\multicolumn{1}{c}{} & \multicolumn{3}{c}{\textbf{Jackson et al. (2021)}} &
\multicolumn{2}{c}{\textbf{sc2ts (wide ARG)}} \\ \hline
\textbf{Sample} &
\textbf{Group} & \textbf{Parents} & \thead{Breakpoint \\
interval(s)} &
\textbf{Parents} & \thead{Breakpoint \\ interval(s)} \\
\hline ALDP-11CF93B & A
    & \thead{B.1.177 \\ Alpha} & 21,255–21,613 & \thead{B.1.177.18 \\ Alpha} &
21,256–22,228 \\
    ALDP-125C4D7 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 &
    \thead{B.1.177.18 \\ Alpha} & 21,256–22,228 \\
ALDP-130BB95 & A & \thead{B.1.177 \\ Alpha} & 21,255–21,613 &
    \thead{B.1.177.18 \\ Alpha} & 21,256–22,228 \\
LIVE-DFCFFE & A & \thead{B.1.177 \\ Alpha} & 18,998–20,294 &
    \thead{B.1.177.18 \\ Alpha} & 21,256–22,228 \\
QEUH-CCCB30 & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 &
    \thead{B.1.36 \\ Alpha} & 6,529–6,955 \\
QEUH-CD0F1F & B & \thead{B.1.36.28 \\ Alpha} & 6,528–6,953 &
    \thead{B.1.36 \\ Alpha} & 6,529–6,955 \\
MILK-1166F52 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 &
    \thead{Alpha \\ B.1.221} & 25,997–27,973 \\
MILK-11C95A6 & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 &
    \thead{Alpha \\ B.1.221} & 25,997–27,973 \\
QEUH-109B25C & C & \thead{Alpha \\ B.1.221.1} & 25,996–27,441 &
    \thead{Alpha \\ B.1.221} & 25,997–27,973 \\
MILK-126FE1F & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 &
    \thead{B.1.36.39 \\ Alpha} & 22,445–23,064 \\
RAND-12671E1 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 &
    \thead{B.1.36.39 \\ Alpha} & 22,445–23,064 \\
RAND-128FA33 & D & \thead{B.1.36.39 \\ Alpha} & 20,703–23,062 &
    \thead{B.1.36.39 \\ Alpha} & 22,445–23,064 \\
CAMC-CBA018 & n/a & \thead{B.1.177 \\ Alpha} & 20,389–21,254 &
    \thead{B.1.177 \\ Alpha} & 17,616--21,256 \\
CAMC-CB7AB3 & n/a & \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3,267–4,474 \\ 20,389–21,254} &
     \thead{Alpha \\ B.1.177 \\ Alpha} & \thead{3268--5389\\17,616--21,256} \\
MILK-103C712 & n/a & \thead{B.1.177.17 \\ Alpha} & \thead{408–444 \\ 26,801–27,876} &
    n/a & n/a \\
QEUH-1067DEF & n/a & \thead{Alpha \\ B.1.177.9} & 10,523–10,869 &
    \thead{Alpha \\ B.1.177} & 7729--10871 \\ \hline
\end{tabular}
\caption{\label{tab:jackson_supplement}Recombinant sequences involving the Alpha (B.1.1.7) variant reported by \cite{Jackson2021-ik} have recombinant ancestry in the wide ARG. The breakpoint intervals and Pango lineage assignments of the parents were taken from Table 2 (3SEQ results) of Jackson et al., except the Pango lineage assignment of the parents of group B recombinants, which were taken from Table 1 (motif-based results).}
\end{table}

@@

\begin{sidewaystable}
% \begin{footnotesize}
\centering
\begin{tabular}{p{1cm}p{1.2cm}p{4.2cm}lll}
\toprule
Focal Pango  & Num origins & Num focal samples (further split by origin, *=nested) & Official Pango parents & \textbf{Main clade}: sc2ts parents & \textbf{Main clade}: other descendants \\
\midrule
\bfseries XA & 1 & \textbf{5} & B.1.1.7 + B.1.177 & B.1.1.7 + B.1.177.18 &  \\
\bfseries XF & 1 & \textbf{2} & B.1.617.2* + BA.1* & AY.4 + BA.1 &  \\
\bfseries XG & 1 & \textbf{32} & BA.1* + BA.2* & BA.1.17 + BA.2 & XAB: 1/48 \\
\bfseries XH & 1 & \textbf{11} & BA.1* + BA.2* & BA.1.20 + BA.2.9 & XAF: 34/35, B.1.1.529: 2, XE: 3/163 \\
\bfseries XK & 1 & \textbf{3} & BA.1* + BA.2* & BA.1.1.1 + BA.2 &  \\
\bfseries XL & 1 & \textbf{10} & BA.1* + BA.2* & BA.1.17.2 + BA.2 & XAB: 1/48, XU: 1/3 \\
\bfseries XR & 1 & \textbf{8} & BA.1.1* + BA.2* & BA.1.1 + BA.2 & XQ: 9/12, XAB: 2/48 \\
\bfseries XS & 1 & \textbf{4} & B.1.617.2* + BA.1.1* & AY.36 + BA.1.1 &  \\
\bfseries XT & 1 & \textbf{1} & BA.1* + BA.2* & BA.2.23 + Unknown & BA.2.23: 1 \\
\bfseries XV & 1 & \textbf{1} & BA.1* + BA.2* & BA.1.1 + BA.2 + Unknown & BA.2: 8 \\
\bfseries XW & 1 & \textbf{11} & BA.1* + BA.2* & BA.1.1.15 + BA.2 & XN: 2/13 \\
\bfseries XY & 1 & \textbf{6} & BA.1* + BA.2* & BA.1.1 + BA.2 & XAF: 1/35 \\
\bfseries XAA & 1 & \textbf{8} & BA.1* + BA.2* & BA.1 + BA.2.9 & XAB: 38/48, XAG: 17/17, XU: 1/3, XQ: 2/12 \\
\bfseries XAC & 1 & \textbf{27} & BA.1* + BA.2* & BA.1.17.2 + BA.2.3 &  \\
\bfseries XAE & 1 & \textbf{9} & BA.1* + BA.2* & BA.1 + BA.2 &  \\
\bfseries XAG & 1 & \textbf{17} & BA.1* + BA.2* & BA.1 + BA.2.9 & XAB: 38/48, XAA: 8/8, XU: 1/3, XQ: 2/12 \\
\bfseries XB & 2 & 57 (\textbf{57}, 1)* & B.1.631 + B.1.634 & B.1 + B.1.627 & B.1.634: 3, B.1.631: 7 \\
\bfseries XC & 2 & 3 (1, \textbf{2}) & AY.29 + B.1.1.7 & AY.103 + B.1.1.7 &  \\
\bfseries XD & 2 & 4 (1, \textbf{3}) & B.1.617.2* + BA.1* & AY.4 + BA.1.15 &  \\
\bfseries XJ & 2 & 2 (1, 1) & BA.1* + BA.2* & No main clade & No main clade \\
\bfseries XAD & 2 & 5 (1, \textbf{4}) & BA.1* + BA.2* & BA.1.1 + BA.2 &  \\
\bfseries XAF & 2 & 35 (\textbf{34}, 1) & BA.1* + BA.2* & BA.1.20 + BA.2.9 & XH: 11/11, B.1.1.529: 2, XE: 3/163 \\
\bfseries XAH & 2 & 12 (\textbf{10}, 2) & BA.1* + BA.2* & BA.1 + BA.2.10 & XZ: 61/66, XAD: 1/5 \\
\bfseries XQ & 3 & 12 (2, \textbf{9}, 1) & BA.1.1* + BA.2* & BA.1.1 + BA.2 & XR: 8/8, XAB: 2/48 \\
\bfseries XU & 3 & 3 (1, 1, 1) & BA.1* + BA.2* & No main clade & No main clade \\
\bfseries XM & 4 & 47 (2, 1, \textbf{40}, 4) & BA.1.1* + BA.2* & BA.1.1 + BA.2 &  \\
\bfseries XN & 4 & 13 (2, 2, \textbf{7}, 2) & BA.1* + BA.2* & BA.1 + BA.2 &  \\
\bfseries XAJ & 4 & 5 (1, 1, \textbf{3}, 1)* & BA.2.12.1* + BA.4* & Unknown & BA.5: 1 \\
\bfseries XE & 5 & 163 (3, \textbf{155}, 1, 3, 2)* & BA.1* + BA.2* & BA.1.17.2 + BA.2 &  \\
\bfseries XZ & 5 & 66 (1, 1, \textbf{61}, 2, 1) & BA.1* + BA.2* & BA.1 + BA.2.10 & XAH: 10/12, XAD: 1/5 \\
\bfseries XAB & 8 & 48 (1, \textbf{38}, 1, 2, 2, 2, 18, 2)* & BA.1* + BA.2* & BA.1 + BA.2.9 & XAA: 8/8, XAG: 17/17, XU: 1/3, XQ: 2/12 \\
\bottomrule
\end{tabular}
\caption{\label{tab:pango-single-origin}
Summary of the Pango X-lineages in the long ARG (excluding XP and XAK whose samples are entirely filtered out, see text). In cases of multiple origins, most X-lineages have a single ``main clade'' (in bold).
Sc2ts inferred parents for the main clade are based on imputed NextClade Pango lineages. Official parents
are taken from Pango designation alias key (TODO: explain the asterisks). Other inferred samples within the main clade are listed, grouped by NextClade Pango designation and clade count / total ARG count (the latter being omitted for non-recombinant designations).
}
\end{sidewaystable}

\begin{table}
\small
\centering
\begin{tabular}{lp{0.2cm}p{0.8cm}lp{0.9cm}ll}
\toprule
Date & $n$ & Causal Pango & Causal strain & Causal \#muts & Sc2ts parents & Breakpoint \\
\midrule
2022-03-24 & 2 & XN & Germany/BY-RKI-I-674341/2022 & 3 & Unknown + BA.2 & (2833,9867) \\
2022-03-29 & 2 & XN & Spain/AB-HGUA-26086757/2022 & 2 & BA.1.1 + BA.2.9 & (2833,4185) \\
2022-04-23 & 7 & XN & Spain/MD-HRYC-12296321/2022 & 2 & BA.1 + BA.2 & (4185,4322) \\
2022-05-17 & 2 & XN & USA/TX-HMH-MCoV-97903/2022 & 5 & BA.2 + BA.2 & (22776,22787) \\
\bottomrule
\end{tabular}
\caption{\label{tab:multiple_origins_table}
Summary of inferred multiple ``origins'' for Nextclade-designated XN samples after filtering
(Official: BA.1* \& BA.2*, breakpoint between 2834 \& 4183, defined from UK strains, earliest
on 2022-02-01;  see \protect\url{https://github.com/cov-lineages/pango-designation/issues/480}).
XN samples are grouped into rows based on their most recent ancestral recombination node;
Details are also provided for the ``causal strain'' that led to the recombination
being proposed by \texttt{sc2ts}. Note that a large number of XN samples were previously removed by filtering. When undersampling causes three or more mutations to lie immediately above the causal strain, \texttt{sc2ts} may mistakenly propose multiple independent origins rather than one.
}
\end{table}
